Evaluating Sound Similarity Metrics for Differentiable, Iterative Sound-Matching

### 2025-09-21 sections/methodology_and_analysis.tex
Updates from Overleaf

\section{Experimental Setup}
\label{sec:experiment_setup}
We systematically evaluate iterative sound-matching pipelines that combine
different differentiable loss functions with differentiable synthesizers.
By ``performance'' we mean the similarity of the synthesized output to a
target, assessed via P-Loss, MSS, and---most importantly---manual listening
tests. This setup allows us to test our central hypothesis: that the relative
performance of a loss function is contingent on the synthesis method, and
thus no universal ``best'' similarity measure exists.

We evaluate four loss functions (Section~\ref{sec:loss_implementation})
across four differentiable synthesizers (Section~\ref{sec:programs}). For
each loss–synthesizer pair, we run a large number of trials and evaluate
their outcomes automatically and manually
(Section~\ref{sec:evaluation_manual_auto}). ``Experiments'' in this context
refer to one complete iterative sound-matching run, from random parameter
initialization through gradient-based optimization until termination. Scores
from these experiments are then aggregated and statistically compared to
determine best-performing losses and synth–loss pairings.

@@

For each program, we provide the Faust code~\cite{orlarey2009faust}, which can be run in the online IDE~\footnote{\url{https://faustide.grame.fr/}}. Faust is a functional language for audio synthesis that can succinctly define signal processing chains. Following Braun's methodology~\cite{braun2024dac}, programs are first defined in Faust, then converted (or \textit{transpiled}) to differentiable Jax functions using the DawDreamer library\footnote{\url{https://github.com/DBraun/DawDreamer}}. 
\begin{lstlisting}[caption={\BPNoise}, label={lst:program0}, language=Faust,
                  float, floatplacement=!H, xleftmargin=1em, xrightmargin=0.5em, firstnumber=0, aboveskip=0em, belowskip=-1em]
import("stdfaust.lib");
lp_cut = hslider("lp_cut",900,50,1000,1);
hp_cut = hslider("hp_cut",100,1,120,1);
process = no.noise:fi.lowpass(3,lp_cut):fi.highpass(10,hp_cut);
\end{lstlisting}

@@

\input{extra_figures/evaluation_graph.tex}
 A post-hoc test is conducted in two stages in order to determine whether there is a difference in performance between loss functions for each program, and if so, which are the best performers. The first stage determines whether there is a difference in group means, with the null hypothesis being that all groups have similar mean ranks. The second stage ranks the loss functions from best to worst using the non-parametric Scott-Knott test (\gls{NPSK})~\cite{tantithamthavorn2017mvt,tantithamthavorn2018optimization}.
The Kruskal-Wallis test is used for the first stage~\cite{kruskal1952use}; this test pools and ranks all evaluation measures for a program, then tests whether the differences in mean rank for all loss function groups is zero. The Kruskal-Wallis test calculates an H statistic that is compared to a chi-square distribution with k-1 degrees of freedom (k is the number of groups, or 4, and degrees of freedom is 3). Using the significance level of 0.05, if the H statistic is greater than the critical value of the chi-square distribution then the null hypothesis is rejected, meaning that at least one group has a mean rank significantly above others~\cite{kruskal1952use}. \todo{More detail for NPSK and leadup to results}

@@

\subsection{Best Performers For Each Program}

Table~\ref{tab:kruskal_auto} shows the Kruskal-Wallis results for every program and evaluation method, we see a significant differences between loss function performance measures (i.e., the bootstrapped distributions of the evaluation for each loss function) in 11 of the 12 cases, with the exception of  MSS for the \AmpMod{} program.

Based on the bootstrapped distribution of scores, the NPSK algorithm ranks the loss functions from 1 (best) to a maximum of 4 (worst). In cases where the distributions are similar, multiple loss functions can be clustered into the same rank. In Figure~\ref{fig:npsk_all}, we use color-coded violin plots to visualize the best performers for each program. The corresponding colors for each rank are
\colorbox{rank1}{\textcolor{black}{\textbf{1}}} \colorbox{rank2}{\textcolor{white}{\textbf{2}}} \colorbox{rank3}{\textcolor{white}{\textbf{3}}} \colorbox{rank4}{\textcolor{black}{\textcolor{white}{\textbf{4}}}}.

@@

\subsubsection{\AmpMod}
IP-Loss and manual hearing tests yield identical rankings, and MSS results vary for for ranks 2 to 4. It is not surprising that MSS results differ from the hearing tests, as we saw in Table~\ref{tab:kruskal_auto}, MSS showed no significant differences between groups. As shown in Figure~\ref{fig:npsk_p2}, all models selected \DTWEnv~as the best performer. This may be due to \DTWEnv's focus on periodic changes in loudness.

>>>>

\section{Experimental Setup}
\label{sec:experiment_setup}
At a glance, our methodology is to conduct controlled, low-level differentiable experiments—without proxy networks—by pairing four DSP-based synthesizers with four differentiable loss functions. We optimize parameters directly via gradients in order to isolate loss–synthesizer interactions. This controlled setup systematically evaluates the performance of iterative sound-matching pipelines, where “performance” is defined as the similarity of the synthesized output to a target. Similarity is assessed with P-Loss, MSS, and—most importantly—manual listening tests. This design allows us to directly test our central hypothesis: that the effectiveness of a loss function depends on the synthesis method, and that no universal “best” similarity measure exists.

Here we discuss the implementation of the four loss functions (Section~\ref{sec:loss_implementation})
across four differentiable synthesizers (Section~\ref{sec:programs}). For
each loss–synthesizer pair, we run a large number of trials and evaluate
their outcomes automatically and manually
(Section~\ref{sec:evaluation_manual_auto}). ``Experiments'' in this context
refer to one complete iterative sound-matching run, from random parameter
initialization through gradient-based optimization until termination. Scores
from these experiments are then aggregated and statistically compared to
determine best-performing losses and synth–loss pairings.

@@

For each program, we provide the Faust code~\cite{orlarey2009faust}, which can be run in the online IDE~\footnote{\url{https://faustide.grame.fr/}}. Faust is a functional language for audio synthesis that can succinctly define signal processing chains. Following Braun's methodology~\cite{braun2024dac}, programs are first defined in Faust, then converted (or \textit{transpiled}) to differentiable JAX functions using the DawDreamer library\footnote{\url{https://github.com/DBraun/DawDreamer}}. 
\begin{lstlisting}[caption={\BPNoise}, label={lst:program0}, language=Faust,
                  float, floatplacement=!H, xleftmargin=1em, xrightmargin=0.5em, firstnumber=0, aboveskip=0em, belowskip=-1em]
import("stdfaust.lib");
lp_cut = hslider("lp_cut",900,50,1000,1);
hp_cut = hslider("hp_cut",100,1,120,1);
process = no.noise:fi.lowpass(3,lp_cut):fi.highpass(10,hp_cut);
\end{lstlisting}

@@

\input{extra_figures/evaluation_graph.tex}
 A post-hoc test is conducted in two stages in order to determine whether there is a difference in performance between loss functions for each program, and if so, which are the best performers. The first stage determines whether there is a difference in group means, with the null hypothesis being that all groups have similar mean ranks. The second stage ranks the loss functions from best to worst using the non-parametric Scott-Knott test (\gls{NPSK})~\cite{tantithamthavorn2017mvt,tantithamthavorn2018optimization}.
The Kruskal-Wallis test is used for the first stage~\cite{kruskal1952use}; this test pools and ranks all evaluation measures for a program, then tests whether the differences in mean rank for all loss function groups is zero. The Kruskal-Wallis test calculates an H statistic that is compared to a chi-square distribution with k-1 degrees of freedom (k is the number of groups, or 4, and degrees of freedom is 3). Using the significance level of 0.05, if the H statistic is greater than the critical value of the chi-square distribution then the null hypothesis is rejected, meaning that at least one group has a mean rank significantly above others~\cite{kruskal1952use}.

@@

\subsection{Best Performers For Each Program}
Table~\ref{tab:kruskal_auto} shows the Kruskal-Wallis results for every program and evaluation method, we see significant differences between loss function performance measures (i.e., the bootstrapped distributions of the evaluation for each loss function) in 11 of the 12 cases, with the exception of  MSS for the \AmpMod{} program.

Based on the bootstrapped distribution of the scores, the NPSK algorithm ranks the loss functions from 1 (best) to a maximum of 4 (worst). In cases where the distributions are similar, multiple loss functions can be clustered into the same rank.

In Figure~\ref{fig:npsk_all}, we use color-coded violin plots to visualize the best performers for each program. The corresponding colors for each rank are
\colorbox{rank1}{\textcolor{black}{\textbf{1}}} \colorbox{rank2}{\textcolor{white}{\textbf{2}}} \colorbox{rank3}{\textcolor{white}{\textbf{3}}} \colorbox{rank4}{\textcolor{black}{\textcolor{white}{\textbf{4}}}}.

@@

\subsubsection{\AmpMod}
P-Loss and manual hearing tests yield identical rankings, and MSS results vary for ranks 2 to 4. It is not surprising that MSS results differ from the hearing tests, as we saw in Table~\ref{tab:kruskal_auto}, MSS showed no significant differences between groups. As shown in Figure~\ref{fig:npsk_p2}, all models selected \DTWEnv~as the best performer. This may be due to \DTWEnv's focus on periodic changes in loudness.

### 2025-09-21 sections/summary_weakness_conclusion.tex
Updates from Overleaf

\subsection{Takeaways}
Perhaps the most important takeaway regarding our main hypothesis is that the loss function that yields the best sound-matching outcomes varies depending on the synthesizer program. This is likely due to the interaction between how the parameters of the synthesizer influence the sound, and the core sonic features used by the loss function.

\textbf{Q1}: To what extent do automatic evaluation metrics agree with manual listening tests? we see somewhat consistent results between the rankings assigned by manual hearing tests (which we take as the ground truth), and automatic measures of P-Loss and MSS. With the exception of P-Loss in \BPNoise{} narrowly selecting a different spectrogram based loss, all measures of performance selected the same top performer. Due to the simplicity of our programs and the occasional deviations, we cannot conclude that these automatic measures are substitutes for human hearing tests. As many previous works have done\todo{cite}, the use of MSS and/or P-Loss as general loss functions does seem appropriate, however, we advise the use of listening tests or custom made loss functions for more specific analysis.

\textbf{Q2}: Are DTW and SIMSE effective measures of loss? If so,

 \DTWEnv{} consistently outperformed other measures in amplitude-modulated synthesis, where envelope alignment is critical. \SIMSESpec{} performed best in subtractive synthesis, where scale-invariance captures noise filtering effects more robustly than L1-based measures. These results indicate that the advantages of novel losses are context-dependent, and the necessity of further creative approaches to differentiable sound-similarity measures.

\textbf{Q3}: Can iterative differentiable optimization be applied effectively in synthesizers using classical DSP functions?
The approach of designing DSP functions in Faust and transpiling to differentiable Faust code yielded meaningful outputs and enabled comparative evaluation of losses. The use of Faust for synthesizer definition was convenient, and transpiling to Jax enabled quick gradient optimization. We were able to run our 1200 experiments within 72 hours on a laptop without a GPU\footnote{Lenovo ThinkPad T480, i5-8350U CPU at 1.70GHz, 32 GB RAM.}.

\subsection{Practical Recommendations:}
Based on our findings, several actionable guidelines emerge. These heuristics can guide practitioners in selecting appropriate similarity measures and synthesis methods for future sound-matching experiments.
\begin{itemize}
    \item In leu of listening tests, MSS or P-Loss are generally useful for large-scale benchmarking, but confirm conclusions with manual listening when accuracy in isolated cases is important. 
    \item For amplitude-modulated synthesis, DTW-based losses (e.g., \DTWEnv{}) are recommended over spectrogram based and JTFS. 
    \item For subtractive/noise-filtered synthesis, spectrogram based losses are most effective. Based on our results, SIMSE appears to be the better method of spectrogram comparisons relative to L1, but the amplitude agnosticism of SIMSE likely plays a role in its poor performance in additive synthesis.  
    \item In our amplitude modulation experiments, \JTFS{} did not outperform other measures, contradicting prior claims of its superiority for mesostructures~\cite{vahidi2023mesostructures}. Thus, our results call into question its effectiveness as a general mesostructure measure
    \item Iterative differentiable optimization via Faust-to-JAX is a viable strategy for defining various differentiable DSP functions, requiring only modest hardware resource and providing a more natural approach to synthesizer definition.
\end{itemize}

>>>>

\subsection{Key Findings}
The most important takeaway regarding our main hypothesis is that the loss function that yields the best sound-matching outcomes varies depending on the synthesizer program. This is likely due to the interaction between how the parameters of the synthesizer influence the sound, and the core sonic features used by the loss function.

\textbf{Q1}: To what extent do automatic evaluation metrics agree with manual listening tests? 
\\We see somewhat consistent results between the rankings assigned by manual hearing tests (which we take as the ground truth), and automatic measures of P-Loss and MSS. With the exception of P-Loss in \BPNoise{} narrowly selecting a different spectrogram based loss, all measures of performance selected the same top performer. However, given the simplicity of our programs and the occasional deviations, we cannot conclude that these automatic measures are substitutes for human hearing tests. As many previous works have done\todo{cite}, the use of MSS and/or P-Loss as general loss functions does seem appropriate, however, we advise the use of listening tests or custom made loss functions for more specific analysis.

\textbf{Q2}: Are DTW and SIMSE effective measures of loss? If so, in what contexts?\\
\DTWEnv{} consistently outperformed other measures in amplitude-modulated synthesis, where envelope alignment is critical. \SIMSESpec{} performed best in subtractive synthesis, where scale-invariance captures noise-filtering effects more robustly than L1-based measures. These results suggest that the advantages of these novel losses are context-dependent, underscoring the need for further exploration of creative differentiable similarity measures.

\textbf{Q3}: Can iterative differentiable optimization be applied effectively in synthesizers using classical DSP functions? \\
The approach of designing DSP functions in Faust and transpiling to differentiable Faust code yielded meaningful outputs and enabled comparative evaluation of losses. The use of Faust for synthesizer definition was convenient, and transpiling to Jax enabled quick gradient optimization. We were able to run our 1200 experiments within 72 hours on a laptop without a GPU\footnote{Lenovo ThinkPad T480, i5-8350U CPU at 1.70GHz, 32 GB RAM.}.

\subsection{Practical Recommendations:}
Based on our findings, several actionable guidelines emerge. These heuristics can guide practitioners in selecting appropriate similarity measures and synthesis methods for future sound-matching experiments.
\begin{itemize}
    \item In lieu of listening tests, MSS or P-Loss are generally useful for large-scale benchmarking, but confirm conclusions with manual listening when accuracy in isolated cases is important. 
    \item For amplitude-modulated synthesis, DTW-based losses (e.g., \DTWEnv{}) are recommended over spectrogram based and JTFS. 
    \item For subtractive/noise-filtered synthesis, spectrogram based losses are most effective. Based on our results, SIMSE appears to be the better method of spectrogram comparisons relative to L1, but the amplitude agnosticism of SIMSE likely plays a role in its poor performance in additive synthesis.  
    \item In our amplitude modulation experiments, \JTFS{} did not outperform other measures, contradicting prior claims of its superiority for mesostructures~\cite{vahidi2023mesostructures}. Thus, our results call into question its effectiveness as a general mesostructure measure
    \item Iterative differentiable optimization via Faust-to-JAX is a viable strategy for defining various differentiable DSP functions, requiring only modest hardware resource and providing a more natural approach to synthesizer definition.
\end{itemize}
